Interpolants and Error Bounds for Modeling and Predicting Variability in Computer Systems

### 2020-04-14 3-Disseration/apps.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

\RequirePackage{fix-cm}
\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
\smartqed  % flush right qed marks, e.g. at end of proof

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.

%
% please place your own definitions here and don't use \def but
% \newcommand{}{}

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\newcolumntype{d}[1]{D{.}{.}{#1}}

%% Create a ``Lemma'' environment that does not have italics font.
\spnewtheorem{plainlemma}{Lemma}{\bf}{}
%% Create a ``Theorem'' environment that is not numbered.
\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
%% Create a ``Proof'' environment that starts with 'Proof.'
\spnewtheorem*{proofdot}{Proof.}{\it}{}

%% \newcommand{\Bigl}{\mathopen\Big}
%% \newcommand{\Bigr}{\mathclose\Big}
%
% Insert the name of "your journal" with
\journalname{Numerical Algorithms}
%
\begin{document}

\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

\author{Thomas C. H. Lux   \and
  Layne T. Watson          \and
  Tyler H. Chang           \and
  Yili Hong                \and
  Kirk Cameron
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Thomas Lux \at
  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
  Blacksburg, VA 24061 \\
  \email{tchlux@vt.edu}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Increases in the quantity of available data have allowed all fields of
science to generate more accurate models of multivariate phenomena.
Regression and interpolation become challenging when the dimension of
data is large, especially while maintaining tractable computational
complexity. Regression is a popular approach to solving approximation
problems with high dimension, however there are often advantages to
interpolation. This paper presents a novel and insightful error bound
for (piecewise) linear interpolation in arbitrary dimension and
contrasts the performance of some interpolation techniques with
popular regression techniques. Empirical results demonstrate the
viability of interpolation for moderately high dimensional
approximation problems, and encourage broader application of
interpolants to multivariate approximation in science.

\keywords{approximation \and regression \and interpolation \and high
  dimension \and error bound}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

%     Introduction     
%======================

%% ===================================================================
\section{Introduction}
\label{sec:introduction}

@@

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are absent from high
dimensional problems. This work extends some known results regarding
the secant method \cite{dennis1996numerical} to construct an
interpolation error bound for problems of arbitrary dimension. These
error bounds are useful, considering the same cannot be said for
regression algorithms in general. The maximum complexity of an
interpolant is bounded by the amount of data available, while the
maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is bounded.

@@

The remainder of this paper is organized as follows. Sections
\ref{sec:regression} and \ref{sec:interpolation} present the
multivariate models. Section \ref{sec:error} gives the error measuring
methodology that is used for collecting empirical results. Section
\ref{sec:theory} contains the theoretical error bounds for
interpolation. Section \ref{sec:data} presents the data sets for
empirical approximation analysis and the approximation results for all
models on these data sets. Section \ref{sec:discussion} analyzes and
discusses the results, their implications, and their
limitations. Finally, Section \ref{sec:conclusion} concludes.

%% ===================================================================
\section{Multivariate Regression}
\label{sec:regression}
Multivariate regressors are capable of accurately modeling a complex
dependence of a response (in $Y$) on multiple variables (represented
as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
minimize some error measure related to data samples
$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
sum of squared differences between modeled function values and true
function values as an error measure. In this section and the next,
some techniques are limited to approximating real valued functions ($Y
\subset \mathbb{R}$). These techniques can be extended to real
vector-valued ranges by repeating the construction for each component
of the vector output. Throughout the following, $x$ denotes a
$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
$d$-tuple data point. Different symbols are used to represent the
approximation function $\hat f$.

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
$B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
maximum number of underlying basis functions. This paper uses an
implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
throughout all experiments.

\subsection{Support Vector Regressor}
\label{sec:svr}
Support vector machines are a common method used in machine learning
classification tasks that can be adapted for the purpose of regression
\cite{basak2007support}. How to build a support vector regressor (SVR)
is beyond the scope of this summary, but the resulting functional fit
$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
 $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
$m$ being determined by the minimization convergence criterion. This
paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
kernel function.

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and five
thousand epochs of SGD for training.

%% ===================================================================
\section{Multivariate Interpolation}
\label{sec:interpolation}
The following interpolation techniques demonstrate a reasonable
variety of approaches to interpolation. All of the presented
interpolants produce approximations that are continuous in value,
which is often a desirable property in applied approximation problems.

\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay triangulation is a well-studied geometric technique for
producing an interpolant \cite{lee1980two}. The Delaunay triangulation
of a set of data points into simplices is such that there are no data
points inside the sphere defined by the vertices of each simplex. For
a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
$\ldots$, $d$, $y \in S$ is a unique convex combination of the
vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
 $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
computational complexity of Delaunay interpolation (for the
implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
d^3)$. Given pathological data the entire triangulation could be
computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
the present application, a Delaunay simplex $S$ containing $y$ is
found, then the $d+1$ vertices (points in $X$) of $S$ are used to
assign weights to each vertex and produce the predicted function value
for point $y$.

\subsection{Modified Shepard}
\label{sec:modified-shepard}

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
$x^{(k)}$. The computational complexity of ShepMod is
$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
is a local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

\subsection{Box Splines}
\label{sec:box-splines}

The box spline model used here is an interpolation technique built
from overlapping box splines \cite{de2013box}. The box splines serve
as basis functions that are shifted and scaled to have support over
box shaped regions. The boxes are constructed in a way to guarantee a
covering of the domain \cite{lux2018novel}. Given a set of box splines
$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
in \cite{lux2018novel} and anchored at interior points
$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

\subsection{Voronoi}
\label{sec:voronoi}

The nearest neighbor algorithm \cite{cover1967nearest} is a
well-studied technique for classification and approximation. Nearest
neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
cell \cite{dirichlet1850reduction}) consisting of all points closer to
$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
the nearest neighbor approximation by utilizing the Voronoi cells to
define support via a generic basis function $v: \mathbb{R}^d
\rightarrow \mathbb{R}_+$ given by
 $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
 {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
has local support, giving the interpolated value
 $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
the approximations are an interpolant involving a convex combination
of known function values $f\bigl(x^{(i)}\bigr)$.

%% ===================================================================
\section{Measuring Error}
\label{sec:error}

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen. The Delaunay
    method chose three source distributions (dotted lines) and
    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
    weighted sum of the three known CDFs produces the predicted CDF
    (dashed line). The KS Statistic (arrow) computed between the true
    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
    example. The KS test null hypothesis is rejected at $p$-value
    0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

When the range of an approximation is the real numbers, error is
reported with summary statistics including: min absolute error, max
absolute error, and absolute error quartiles. When the range of an
approximation is the space of cumulative distribution functions, the
Kolmogorov-Smirnov statistic (max-norm difference between the
functions) is used.

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}.

The performance of approximation techniques that predict probability
functions can be analyzed through a variety of summary
statistics. This work uses the max absolute difference, also known as
the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing
two CDFs while only assuming stationarity, finite mean, and finite
variance. The null hypothesis (that two CDFs come from the same
underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the round-trip prediction methodology from known and
predicted distributions to the calculation of error can be seen in
Figure \ref{fig:prediction-example}.

\section{Theoretical Error Bound}
\label{sec:theory}

This section presents the theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
paper compute predictions as convex combinations of observed function
values, which may allow for straight forward extensions of this error
bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
  \rightarrow \mathbb{R}$, and $\nabla f \in
  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
  continuous functions in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}

\begin{proofdot}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
  in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
       &= \big|g(1) - g(0) - g'(0) \big| & \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
       &= \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
  \qed
\end{proofdot}

\begin{plainlemma}
  \label{lemma:2}
  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
  where $\sigma_d$ is the smallest singular value of $M.$
\end{plainlemma}

\begin{proofdot}
  Using the facts that $M$ and $M^t$ have the same singular values,
  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
  \begin{align*}
    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
  \end{align*}
  \qed
\end{proofdot}

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
  \big\rangle = 0.$ Now
  \begin{align*}
    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
  \end{align*}
  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
  \qed
\end{proofdot}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error. In light of this error bound, an empirical evaluation of
presented algorithms follows.

%% ===================================================================
\vspace{-2mm}
\section{Data and Empirical Analysis}
\label{sec:data}

The theoretical results constructed in Section \ref{sec:theory} for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, this paper utilizes five
different data sets of varying dimension and application to construct
approximations and compare the accuracy of different techniques.

In the following five subsections the sources and targets of each test
data set are described, as well as challenges and limitations related
to approximating these data. The distribution of response values being
modeled is presented followed by the distribution of approximation
errors for each algorithm. The plots for all five data sets have the
same format.

All five data sets are rescaled such that the domain of approximation
is the unit hypercube. The range of the first four data sets is the
real numbers, while the range of the fifth data set is the space of
cumulative distribution functions. All approximation techniques are
applied to the first four data sets, while only those interpolants
whose approximations are convex combinations of observed data are
applied to the final data set.

All approximations are constructed using $k$-fold cross validation as
described in \cite{kohavi1995study} with $k=10$. This approach
randomly partitions data into $k$ (nearly) equal sized sets. Each
algorithm is then evaluated by constructing an approximation over each
unique union of $k-1$ elements of the partition, making predictions
for points in the remaining element. As a result, each observed data
point is used in the construction of $k-1$ different approximations
and is approximated exactly once. The $k$-fold cross validation method
is data-efficient and provides an unbiased estimate of the expected
prediction error \cite{kohavi1995study}, however it should be noted
that neither this method nor others can provide a universally unbiased
estimator for the variance of prediction error \cite{bengio2004no}.

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results are located in the
Appendix. All of the test data sets capture underlying functions that
are almost certainly stochastic. As described in Section
\ref{sec:introduction}, regression techniques appear most appropriate
for these problems. However, typically data grows exponentially more
sparse with increasing dimension. Given sparse data, regressors tend
towards interpolation.

%% -------------------------------------------------------------------
\subsection{Forest Fire ($n = 504$, $d = 12$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

>>>>

Computational variability presents itself in a variety of forms. Processes that are apparently identical in a cloud computing or high performance computing (HPC) environment may take different amounts of time to complete the same job. This variability can cause unintentional violations of service level agreements in cloud computing applications or indicate suboptimal performance in HPC applications. The sources of variability, however, are distributed throughout the system stack and often difficult to identify. The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance measures without any prior knowledge of system architecture. Some examples of interesting performance measures that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other numeric measure of performance.

Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks. In this work, modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.

Many previous works attempting to model system performance have used simulated environments \cite{grobelny2007fase,wang2009simulation,wang2013towards}. \citet{grobelny2007fase} refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system. Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system. \cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}. These limited statistical models have provided satisfactory performance in narrow application settings. However these techniques require additional code annotations, hardware abstractions, or additional assumptions about the behavior of the application in order to generate models. In contrast, the approaches that are presented here require no modifications of applications, no architectural abstractions, nor any structural descriptions of the input data being modeled. The techniques used in this work only require performance data as input.

Most existing work on performance variability has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}. Yet, system I/O variability has been particularly difficult for statistical models to capture \cite{bailey2005performance}. The prior work attempting to model I/O variability has been similarly limited to one or two system parameters \cite{lofstead2010managing}.

Chapters \ref{ch:naive} and \ref{ch:strong} present and evaluate applications of multivariate interpolation to the domain of computer system I/O throughput. Interpolants and regressors are used to predict the different measures of variability (e.g., mean, variance, cumulative distribution functions) for the expected I/O throughput on a system with previously unseen configurations. Beyond these I/O case studies, the techniques discussed can tractably model tens of interacting system parameters with tens of thousands of known configurations. A major contribution of this work is a modeling framework that uses multivariate interpolation to capture precise characteristics (via cumulative distribution functions) of arbitrary performance measure on any type of computer system.

\section{Broader Applications of Approximation}

@@

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are absent from high
dimensional problems. Chapter \ref{ch:error} extends some known
results regarding the secant method \cite{dennis1996numerical} to
construct an interpolation error bound for problems of arbitrary
dimension. These error bounds are useful, considering the same cannot
be said for regression algorithms in general. The maximum complexity
of an interpolant is bounded by the amount of data available, while
the maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is
bounded. These generic bounds imply that interpolants may be suitable
for a broader class of approximation problems than (heuristically
chosen) regression techniques.

@@

\section{An Initial Application of Approximation}

Performance tuning is often an experimentally complex and time
intensive chore necessary for configuring High Performance Computing
(HPC) systems. The procedures for this tuning vary largely from system
to system and are often subjectively guided by the system
engineer(s). Once a desired level of performance is achieved, an HPC
system may only be incrementally reconfigured as required by updates
or specific jobs. In the case that a system has changing workloads or
nonstationary performance objectives that range from maximizing
computational throughput to minimizing power consumption and system
variability, it becomes clear that a more effective and automated tool
is needed for configuring systems. This scenario presents a
challenging and important application of multivariate approximation
and interpolation techniques.

Predicting the performance of an HPC system is a challenging problem
that is primarily attempted in one of two ways: (1) build a
statistical model of the performance by running experiments on the
system at select settings, or (2) run artificial experiments using a
simplified simulation of the target system to estimate architecture
and application bottlenecks. In this paper the proposed multivariate
modeling techniques rest in the first category, and they represent a
notable increase in the ability to model complex interactions between
system parameters.

Many previous works attempting to model system performance have used
simulated environments to estimate the performance of a system
\cite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
of these works refer to statistical models as being oversimplified and
not capable of capturing the true complexity of the underlying
system. This claim is partially correct, noting that a large portion
of predictive statistical models rely on simplifying the model to one
or two parameters
\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
These limited statistical models have provided satisfactory
performance in very narrow application settings. Many of the
aforementioned statistical modeling techniques claim to generalize,
while simultaneously requiring additional code annotations, hardware
abstractions, or additional application level understandings in order
to generate models. The approach presented here requires no
modifications of the application, no architectural abstractions, nor
any structural descriptions of the input data being modeled. The
techniques used are purely mathematical and only need performance data
as input.

Among the statistical models presented in prior works,
\cite{bailey2005performance} specifically mention that it is difficult
for the simplified models to capture variability introduced by
I/O. System variability in general has become a problem of increasing
interest to the HPC and systems communities, however most of the work
has focused on operating system (OS) induced variability
\cite{beckman2008benchmarking,de2007identifying}. The work that has
focused on managing I/O variability does not use any sophisticated
modeling techniques \cite{lofstead2010managing}. Hence, Chapter
\ref{ch:naive} presents a case study applying advanced mathematical
and statistical modeling techniques to the domain of HPC I/O
characteristics. The models are used to predict the mean throughput of
a system and the variance in throughput of a system. The discussion
section that follows outlines how the techniques presented can be
applied to any performance metric and any system.
